Synthesising Complex Termination Arguments

### 2014-06-23 papers/termination/paper.tex
2nd order  termination

\section{Motivational Examples} \label{sec:motivation}
In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
Now, we provide some simple examples that broke those assumptions and

We next discus some intricate examples that prove challenging for existent termination analysis.
%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
%frequently in practise and are not properly handled by existing techniques.

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure} \\

\hline

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while ( q >= 0 ) {
    q = q - y;
    y = y + 1;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.d}
\end{subfigure}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\caption{}
 \label{fig:motivation.f}
\end{subfigure}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}
\caption{}
 \label{fig:motivation.g}
\end{subfigure} \\
\hline

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

\end{tabular}
\caption{Sample programs}\label{fig:motivation}
\end{figure*}

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.  We will make use of this observation to frame the halting
problem as a program synthesis problem, allowing us to employ the program synthesiser described in
Section~\ref{sec:synthesis}.

@@

With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
above is an $\exists \forall$ formula where the the universally quantified variables only range over
ground terms.  We refer to this syntactic class as the \emph{synthesis fragment} of second-order
logic, because program synthesis problem can be described as finding a satisfying assigment to the
following formula:

>>>>

\section{Motivational Examples} \label{sec:motivation}
We next discus some intricate examples that prove challenging for existent termination analysis.
%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
%frequently in practise and are not properly handled by existing techniques.

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure}%

\end{tabular}
\end{figure*}

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.  This fragment
consists of $\exists \forall$ formulae where the the universally quantified variables only range over
ground terms and has close ties to the program synthesis problem.  We will make use of this observation
to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
Section~\ref{sec:synthesis}.

@@

With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
described as finding a satisfying assigment to the following formula, which is the general case of
the synthesis fragment:
